A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-21 paper.tex
Change to refer to palaramara for "The" ARG
Drop "in retrospect" para
Closes #308

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer complete eARGs (section XXX),
which contain specific causal recombination \emph{events}
for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer causal events but rather
infer gARGs with differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
Therefore, they contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

% \relate\ "postorder" first inferring local topologies and then identifies edges
% that persist from one tree, while
% local trees only emerge from \tsinfer\ as a consequence of the copying
% patterns inferred among ancestral haplotypes.

% % More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
% % for discussion. I'm sure there were other examples of people saying
% % that
% (as claimed by e.g.~\citet{hejase2020summary}),
% and information about recombination and the correlation structure
% between trees is not ``all or nothing''.
% Both methods capture correlation structure between trees:
% \relate\ first infers local tree topologies and then identifies edges
% that persist from one tree to the next as a separate step, while
% \texttt{tsinfer} explicitly constructs nodes that are shared across local
% trees.
% % Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
% % explicit recombination events we cannot compute a likelihood
% % under the coalescent with recombination (see appendix XXX).
% % [Some stuff about whether we should see this as necessary or not,
% % but maybe we should keep this for the conclusion?]

@@

% These properties have interesting consequences when we
% consider the inference strategy of esimating local trees
% independenty and subsequently stitching them together
% into an ARG. Two recent methods take this approach,
% \relate\ and
% \texttt{espalier}~\citep{rasmussen2022espalier},
% and have different strategies for reconciling the local
% trees into an ARG. [Relate does something fairly lightweight,
% but by default imprecise.]
% Espalier, on the other hand, is specifically in interested
% in the recombination events and therefore attempts to infer
% the precise ``subtree prune-and-regraft'' (SPR)
% operations~\citep{hein1990reconstructing,song2003on,song2006properties}
% using heuristics~\citep{rasmussen2022espalier}.
% Inferring the SPRs between
% leaf-labelled trees is NP-hard in
% general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

% JK TODO: finish up this section with a quick version of the
% para here. Basically say, people often say an ARG is a
% set of local trees cite cite, but it's not as simple as that as we
% can see from the example (probably).

@@

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity, but see below for confusion around this point)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely}.

@@

In retrospect, the choice of terminology around the succinct
tree sequence data structure was unfortunate, and has led to
significant confusion. As first introduced by~\cite{kelleher2016efficient}
a ``tree sequence`` was defined as the set of ``coalescence records''
that is output by Hudson's simulation algorithm.
This was then generalised to use the
tabular encoding described above and formally described as a ``succinct
tree sequence''~\citep{kelleher2018efficient}.
The ``succinct'' prefix here is intended to connect
with the concept of succinct data structures
\citep[e.g.][]{gog2014theory},
which have near-optimal
space usage but still support efficient retrieval and computation.
The concrete data structures of the succinct tree sequence, precisely
defined and with a particular focus on computational efficiency
did not have an obvious connection to the ARG definitions
provided by Griffiths and
colleagues~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}, and with the lack
of any other precise definitions, it seemed best to avoid the
term to prevent confusion. Unfortunately, however, the opposite
has occured. In describing the output of \tsinfer\ as a ``tree
sequence''~\citep{kelleher2019inferring} many
have---not unreasonably, but incorrectly---concluded that the
output is a collection of \emph{independent}
% MORE?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
trees~\citep[e.g.,][]{hejase2020summary,ignatieva2021kwarg}.
% Could say more here, but maybe we'll have a note about terminology
% in the discussion

@@

[NOTE This is a pedantic and slightly bullshitty to end up on,
just writing it down to see if it works. Should probably cut.]
A final point to note, is that
if we are using the term ARG
to loosely refer to genetic genealogies of varying
levels of detail, either inferred or simulated,
it makes little sense to refer ``the'' Ancestral Recombination
Graph (i.e, in  the singular, like ``the tree'' or
``the phylogeny'')
% TODO add more if we keep this
as favoured by some~\citep[e.g.][]{
arenas2013importance,hubisz2020inference,zhang2023biobank}.
While there is of course a single ``true'' ARG that describes
the actual history of a sample, in reality there will be astronically
many ARGs that are compatible with the data, even under
parsimony conditions or when sampling from a model.
There are many different levels of granularity at which we
can reason about recombination, and many different ARGs.
We should embrace this diversity.

>>>>

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer complete eARGs (section XXX),
which contain specific causal recombination \emph{events}
for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer causal events but rather
infer gARGs with differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
Therefore, they contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.
The output of these programs is certainly not a collection of
\emph{unrelated} local trees, as implied by,
e.g.,
\citet{hejase2020summary} when they state that \tsinfer\
``does not explicitly infer an ARG but rather a sequence of local
gene trees''.

% More cites? Not worth harping on about now?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38

@@

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
independently and subsequently stitching them together
into an ARG.
Two recent methods take this approach,
\relate\ and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and the identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among small samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

% can see from the example (probably).

@@

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely}.

@@

@@

[NOTE This is a pedantic and slightly bullshitty to end up on,
just writing it down to see if it works. Should probably cut.]
A final point to note, is that
if we are using the term ARG
to loosely refer to genetic genealogies of varying
levels of detail, either inferred or simulated,
it makes little sense to refer ``the'' Ancestral Recombination
Graph (i.e, in  the singular, like ``the tree'' or
``the phylogeny'')
% TODO add more if we keep this
as favoured by some~\citep[e.g.][]{
arenas2013importance,palamara2016argon,hubisz2020inference}.
While there is of course a single ``true'' ARG that describes
the actual history of a sample, in reality there will be astronically
many ARGs that are compatible with the data, even under
parsimony conditions or when sampling from a model.
There are many different levels of granularity at which we
can reason about recombination, and many different ARGs.
We should embrace this diversity.
